Synthesising Recursive Functions for First-Order Model Counting:\\ Challenges, Progress, and Conjectures

### 2023-06-08 doc/paper/paper.tex
Camera-ready

\Cref{thm:correctness1} is equivalent to Proposition~3 by Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11}. For the proofs of this and other
theorems, see supplementary material.

@@

Recall that, in the case of \textsc{ForcLift}, solving a (W)FOMC problem
consists of two parts: compilation (via search) and propagation. (A similar
dichotomy applies to \textsc{Crane} as well.) The complexity of the former
depends only on the formula and so far has received very little attention except
for the work by Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15}. The
complexity of the latter, on the other hand, is a function of domain sizes and
can be determined without measuring runtime. We establish the asymptotic
complexity of a solution by counting the number of arithmetic operations needed
to follow the definitions of constituent functions without recomputing the same
quantity multiple times. In particular, we assume that each function call and
binomial coefficient is computed at most once, and computing $\binom{n}{k}$
takes $\Theta(nk)$ time. For example, the complexity of \cref{eq:solution} is
$\Theta(mn)$ since $f(m, n)$ can be computed by a dynamic programming algorithm
that computes $f(i, j)$ for all $i = 0, \dots, m$ and $j = 0, \dots, n$, taking
a constant amount of time on each $f(i, j)$.

Let $m = |\Gamma|$ and $n = |\Delta|$ be domain sizes. We summarise the results
in \cref{tbl:results}, where we compare the solutions found by both algorithms
to the manually-constructed ways of computing the same quantities that have been
proposed before.\footnote{\url{https://oeis.org},
  \url{https://scicomp.stackexchange.com/q/30049}} Examining the last two
columns of the table, we see that the two algorithms perform equally well on
instances that could already be solved by \textsc{ForcLift}. However,
\textsc{Crane} can also solve all but one of the instances that
\textsc{ForcLift} fails on in at most cubic time. See supplementary material for
the exact solutions produced by \textsc{Crane}.

@@

\section{Conclusion and Future Work}\label{sec:conclusion}
\setlength{\multlinegap}{0pt}

This paper presents \textsc{Crane}---an extension of
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} that benefits from a more
general version of domain recursion and support for graphs with cycles. Our
results in \cref{sec:results} reveal a range of counting problems that became
newly feasible as a result of these enhancements, including instances outside of
currently-known domain-liftable fragments such as \Ctwo{}. (Although we focused
on unweighted counting, \textsc{ForcLift}'s support for weights trivially
transfers to \textsc{Crane} too.) The common thread across these newly liftable
problems is a version of partial injectivity. Thus, we formulate the following
conjecture.
\begin{conjecture}
  Let \IFO{} be the class of formulas in first-order logic that contain clauses
  with at most two variables as well as any number of copies of
  \begin{align*}
    &\begin{multlined}[t][0.9\linewidth]
      (\forall X_{1} \in \Delta_{1}\text{. }\dots\,\forall X_{n} \in \Delta_{n}\text{. }\forall Y \in \Delta_{i}\text{. }\predicate(X_{1}, \dots, X_{n}) \land{}\\
      \predicate(X_{1}, \dots, X_{i-1}, Y, X_{i+1}, \dots, X_{n}) \Rightarrow X_{i} = Y) \land{}
    \end{multlined}\\
    &\begin{multlined}[t][0.9\linewidth]
      (\forall X_{1} \in \Delta_{1}\text{. }\dots\,\forall X_{n} \in \Delta_{n}\text{. }\forall Y \in \Delta_{j}\text{. }\predicate(X_{1}, \dots, X_{n}) \land{}\\
      \predicate(X_{1}, \dots, X_{j-1}, Y, X_{j+1}, \dots, X_{n}) \Rightarrow X_{j} = Y)
    \end{multlined}
  \end{align*}
  for some predicate $\predicate/n$, domains
  $\Delta_{1}, \dots, \Delta_{n}$, and $i, j \in \{\, 1, \dots, n \,\}$. Then
  \IFO{} is liftable by \textsc{Crane}.
\end{conjecture}
Since most of the instances in \cref{sec:results} are in \Ctwo{}, we can also
conjecture that \Ctwo{} is liftable by \textsc{Crane}.

The most important direction for future work is to automate the process in
\cref{fig:processcrane}. First, we need a way to find the base cases for the
recursive definitions produced by \textsc{Crane}. Second, since the first
solution found by \textsc{Crane} is not always optimal in terms of its
complexity, an automated way to determine the asymptotic complexity of a
solution would be helpful as well. Third, as we saw at the end of
\cref{example:interpretation}, the functions constructed by \textsc{Crane} can
often benefit from elementary algebraic simplifications. This process can be
automated using a computer algebra system. Furthermore, note that having the
solution to a (W)FOMC problem expressed in terms of functions enables many new
possibilities such as
\begin{enumerate*}[label=(\roman*)]
  \item the use of more sophisticated simplification techniques,
  \item asymptotic analysis of, e.g., how the model count grows as the domain
  size goes to infinity, and
  \item answering questions parameterised by domain sizes, e.g., `how big does
  domain $\Delta$ have to be for the probability of event $E$ to be above
  \SI{95}{\percent}?'
\end{enumerate*}
In addition to the potential impact on areas of artificial intelligence (AI)
like statistical relational AI~\cite{DBLP:series/synthesis/2016Raedt},
\textsc{Crane} could be beneficial to research in combinatorics as
well~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

>>>>

\Cref{thm:correctness1} is equivalent to Proposition~3 by Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11}. For the proofs of this and other
theorems, see the full version of this
paper\footnote{\url{https://arxiv.org/abs/2306.04189}}.

@@

Recall that, in the case of \textsc{ForcLift}, solving a (W)FOMC problem
consists of two parts: compilation (via search) and propagation. (A similar
dichotomy applies to \textsc{Crane} as well.) The complexity of the former
depends only on the formula and so far has received very little attention except
for the work by Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15}. The
complexity of the latter, on the other hand, is a function of domain sizes and
can be determined without measuring runtime. We establish the asymptotic
complexity of a solution by counting the number of arithmetic operations needed
to follow the definitions of constituent functions without recomputing the same
quantity multiple times.\footnote{Although this is not done by default, one
  could also use arbitrary-precision arithmetic and consider bit complexity
  instead.} In particular, we assume that each function call and binomial
coefficient is computed at most once, and computing $\binom{n}{k}$ takes
$\Theta(nk)$ time. For example, the complexity of \cref{eq:solution} is
$\Theta(mn)$ since $f(m, n)$ can be computed by a dynamic programming algorithm
that computes $f(i, j)$ for all $i = 0, \dots, m$ and $j = 0, \dots, n$, taking
a constant amount of time on each $f(i, j)$.

Let $m = |\Gamma|$ and $n = |\Delta|$ be domain sizes. We summarise the results
in \cref{tbl:results}, where we compare the solutions found by both algorithms
to the manually-constructed ways of computing the same quantities that have been
proposed before.\footnote{\url{https://oeis.org},
  \url{https://scicomp.stackexchange.com/q/30049}} Examining the last two
columns of the table, we see that the two algorithms perform equally well on
instances that could already be solved by \textsc{ForcLift}. However,
\textsc{Crane} can also solve all but one of the instances that
\textsc{ForcLift} fails on in at most cubic time. See the full version of this
paper\footnote{\url{https://arxiv.org/abs/2306.04189}} for the exact solutions
produced by \textsc{Crane}.

@@

\section{Discussion}\label{sec:conclusion}
\setlength{\multlinegap}{0pt}

This paper presents \textsc{Crane}---an extension of
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} that benefits from a more
general version of domain recursion and support for graphs with cycles. In
\cref{sec:results}, we listed a range of counting problems that became newly
feasible as a result, including instances outside of currently-known
domain-liftable fragments such as \Ctwo{}. (Although we focused on unweighted
counting, \textsc{ForcLift}'s support for weights trivially transfers to
\textsc{Crane} too.) The common thread across these newly liftable problems is a
version of partial injectivity. Thus, we formulate the following conjecture for
a class of formulas with an injectivity-like constraint on at most two
parameters of a predicate $\predicate$.

\begin{conjecture}\label{conj}
  Let \IFO{} be the class of formulas in first-order logic that contain clauses
  with at most two variables as well as any number of copies of
  \begin{align*}
    &\begin{multlined}[t][0.9\linewidth]
      (\forall X_{1} \in \Delta_{1}\text{. }\dots\,\forall X_{n} \in \Delta_{n}\text{. }\forall Y \in \Delta_{i}\text{. }\predicate(X_{1}, \dots, X_{n}) \land{}\\
      \predicate(X_{1}, \dots, X_{i-1}, Y, X_{i+1}, \dots, X_{n}) \Rightarrow X_{i} = Y) \land{}
    \end{multlined}\\
    &\begin{multlined}[t][0.9\linewidth]
      (\forall X_{1} \in \Delta_{1}\text{. }\dots\,\forall X_{n} \in \Delta_{n}\text{. }\forall Y \in \Delta_{j}\text{. }\predicate(X_{1}, \dots, X_{n}) \land{}\\
      \predicate(X_{1}, \dots, X_{j-1}, Y, X_{j+1}, \dots, X_{n}) \Rightarrow X_{j} = Y)
    \end{multlined}
  \end{align*}
  for some predicate $\predicate/n$, domains
  $\Delta_{1}, \dots, \Delta_{n}$, and $i, j \in \{\, 1, \dots, n \,\}$. Then
  \IFO{} is liftable by \textsc{Crane}.
\end{conjecture}

Note that the unsolved instance in \cref{tbl:results} does not invalidate
\cref{conj} for two reasons. First, in our experiments, the depth of the search
procedure is limited, so there is a possibility that increasing the depth would
make the algorithm succeed. Second, the conjecture assumes distinct domains,
while the unsolved instance pertains to endofunctions.

% In general, our results demonstrate that Crane
% performs better when all domains are distinct. A solution for distinct domains
% can always be applied to the case when some domains are equal.

% ===========================================

% Since most of the instances in \cref{sec:results} are in \Ctwo{}, we can also
% conjecture that \Ctwo{} is liftable by \textsc{Crane}.

% Every sentence in C2 can be rewritten to avoid counting quantifiers by
% introducing more variables. For example, to claim that there are exactly two
% Xs that satisfy p(X), one would introduce two existentially-quantified
% variables X_1 and X_2 such that X_1 != X_2, p(X_1), and p(X_2) and add a
% constraint that any other variable X_3 such that p(X_3) has to be equal to
% either X_1 or X_2.

The most important direction for future work is to automate the process in
\cref{fig:processcrane}. First, we need a way to find the base cases for the
recursive definitions produced by \textsc{Crane}. Second, since the first
solution found by \textsc{Crane} is not always optimal in terms of its
complexity, an automated way to determine the asymptotic complexity of a
solution would be helpful as well. Third, as we saw at the end of
\cref{example:interpretation}, the functions constructed by \textsc{Crane} can
often benefit from elementary algebraic simplifications. This process can be
automated using a computer algebra system. Furthermore, note that having the
solution to a (W)FOMC problem expressed in terms of functions enables many new
possibilities such as
\begin{enumerate*}[label=(\roman*)]
  \item the use of more sophisticated simplification techniques,
  \item asymptotic analysis of, e.g., how the model count grows as the domain
  size goes to infinity, and
  \item answering questions parameterised by domain sizes, e.g., `how big does
  domain $\Delta$ have to be for the probability of event $E$ to be above
  \SI{95}{\percent}?'
\end{enumerate*}

\section*{Acknowledgments}

We thank Anna L.D. Latour and Kuldeep S. Meel for comments on earlier drafts of
this paper. Most of the work for this paper was done while the first author was
a PhD student at the University of Edinburgh. The first author was supported by
the EPSRC Centre for Doctoral Training in Robotics and Autonomous Systems,
funded by the UK Engineering and Physical Sciences Research Council (grant
EP/L016834/1). The second author was partly supported by a Royal Society
University Research Fellowship, UK, and partly supported by a Grant from the
UKRI Strategic Priorities Fund, UK to the UKRI Research Node on Trustworthy
Autonomous Systems Governance and Regulation (EP/ V026607/1, 2020–2024). For the
purpose of open access, the authors have applied a Creative Commons Attribution
(CC BY) licence to any Author Accepted Manuscript version arising from this
submission.
